---
title: "COVID-19 Case Rates Across Canadian Health Regions"
output:
   html_document:
    theme: null
---

## Background

This report was developed to highlight the current COVID-19 situation across Canada. Providing case rate statistics by local health regions is better suited to detect outbreaks or local health concerns. The case rate used here is the _count of new cases reported in the previous 7 day, divided by 100,000 population_.  

This report was completed using RMarkdown and was completed as part of a web scraping, automation, and visualization project. COVID-19 data is automatically scraped, processed, visualized, and uploaded everyday at ____.  

Provincial and territorial COVID-19 data was imported from Canada.ca, and health region data was imported from [COVID-19 Canada](https://resources-covid19canada.hub.arcgis.com). Note that some regions report data at different frequencies, so case rates may be a day or two delayed in a few instances. 

The ggplot2 package was used to display the provincial and territorial case rate, and the reactable package was used to develop the interactive table. 

## Provincial and Territorial COVID-19 Case Rates

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(reactable)
library(htmltools)
library(lubridate)
library(hrbrthemes)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(reshape2)
library(fuzzyjoin)
library(tableHTML)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

data <- read.csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv") ## Provincial and territorial COVID data


######## Select the necessary data.

data$date <- as.Date(data$date)

data <- data %>% select(prname, date, numtoday)  %>% filter(date > (max(date)) - 20) %>%
               filter( prname != "Canada" & prname != "Repatriated travellers")   ## selects the variables of interest, reduced rows to the last 20 days
                                                                                      ## we'll need to calculate  over two different weeks - so provides some padding for different reporting times.



###### Manipulate data to prepare for visualization.. see if I can do a panel chart with two vertical bar charts, one shows the rate change between weeks.

# read in new StatsCan Q3 population sizes.

pops <- read.csv("StatsCanQ3_pop.csv", header = T)
colnames(pops)[1] <- "PT"
pops$POP <- as.numeric(pops$POP)

######### forloop, get sums of case counts per province
#
# PT_df <- c()
# cases_7_df <- c()
# cases_previous_week_df <- c()
# change_df <- c()

table_values <- data.frame()
for (i in data$prname){
  population <- pops[pops$PT == i, 2] /100000
  cases_14 <- data %>% filter(prname == i) %>% filter(date > (max(date) - 14))
  cases_7 <- cases_14 %>% filter(date > (max(date) - 7)) %>% select(numtoday) %>% sum() / population
  cases_previous_week <- cases_14 %>% filter(date < (max(date)-6)) %>% select(numtoday) %>% sum() / population
  change <- cases_7 - cases_previous_week

  PTs <- data$prname[1:13]
  ii <- which(PTs == i)

  table_values[which(PTs == i),1:4] <- data.frame(i, cases_7, cases_previous_week, change)
}


##### plot it
names(table_values)[1] <- "Province"


m_df <- table_values[,-3]
m_df <- melt(m_df, id.vars= 1)
m_df$fill <- ifelse(m_df$value < 0, 3, 2)
m_df$fill[m_df$variable == "cases_7"] <- 1
m_df$fill <- factor(m_df$fill)

m_df$Province[m_df$Province== "Quebec"] <- "Québec"

# Create order facotrs levels to plot largest to smallest values.
province_order <- table_values$Province[order(table_values$cases_7)]
province_order[province_order == "Quebec"] <- "Québec"


m_df$Province <- factor(m_df$Province, levels = province_order, ordered = T)


colors <- c("gray64", "red", "green1")
# New facet label names for dose variable
var.labs <- c("cases_7" = "NEW CASES OVER 7 DAY PERIOD / 100,000 POPULATION",
              "change" = "CHANGE IN CASE RATE FROM PREVIOUS WEEK")


#windowsFonts()

library(extrafont)
# font_import()
# loadfonts(device = "win")

# m_df$value2 <- m_df$value
# m_df$value2[m_df$variable == "change"] <- m_df$value2[m_df$variable == "change"] - min(m_df$value2[m_df$variable == "change"])


#m_df$hjust <- ifelse(m_df$value >= 0, -0.2, -0.1)

m_df$hjust <- ifelse(m_df$value >= 0, -0.3, 1.1)
m_df$vjust <- ifelse(m_df$value >= 0, "top", 0)

pfw <- ggplot(data = m_df, aes(x = Province, y = value, group= variable, fill = fill, hjust = hjust)) +
  geom_bar(stat = "identity", data = m_df, colour ="black") + scale_fill_manual(values=colors)+
  geom_text(aes(label = round(value,1)), colour = "black") +
  facet_grid(~ variable, labeller = labeller(variable = var.labs, size = 10), scales = "free", space = "free_x") +
  guides(fill = FALSE) +  # to remove the legend
  theme_bw() + coord_flip() + labs(x="", y="") + theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  theme(strip.text = element_text(size=10.5)) + theme(strip.background =element_rect(fill="gray96"))+
  theme(axis.text.y = element_text(size = 11, colour = "black"))+ theme(text=element_text(family="TT Times New Roman")) + scale_y_continuous(expand = c(0.1, 4)) + ggtitle("Provincial and Territorial COVID-19 Case Rates") + theme(plot.title = element_text(size=18))
                                                                        # + ylim((min(table_values$change)*1), (max(table_values$cases_7)*1.25))

# change scales...change number position on negative side. 


# ymax_case <- max(m_df$value)
# scales




```


```{r echo=FALSE, message=FALSE, warning=FALSE} 

#ON_data <- read.csv("COVID-19_ON_data.csv")
#QC_data<- read.csv(url("https://msss.gouv.qc.ca/professionnels/statistiques/documents/covid19/COVID19_Qc_RapportINSPQ_VigieCategories.csv"))

#AB_data <- read.csv(url("blob:https://www.alberta.ca/4964468a-4364-4daf-b61b-97a325e7aa1f"))



HZ_data <- read.csv(url("https://opendata.arcgis.com/api/v3/datasets/3aa9f7b1428642998fa399c57dad8045_1/downloads/data?format=csv&spatialRefId=4326"))


### remove time, order date, group by HR, minus previous day. 

HZ_data$Last_Updated <- as.Date(gsub("\\ .*","",HZ_data$Last_Updated))
HZ_data$HR_UID <- as.character(HZ_data$HR_UID)

HZ_data <- HZ_data %>% select(Province, HR_UID, CaseCount, Last_Updated) %>% filter(Last_Updated > max(Last_Updated) - 35)

# HZ_new_cases <- 
  
ii <- 1
HZ_table_values <- data.frame()
for (i in unique(HZ_data$HR_UID)){
  #population <- HZ_data %>% filter(HR_UID == i) %>% select(TotalPop2019) %>% first()/100000
  #df$Date_difference_in_days <-difftime(df$date1,df$date2,units=c("days"))
  cases_14 <- HZ_data %>% filter(HR_UID == i)  %>% filter(Last_Updated > (max(Last_Updated) - 16)) %>% arrange(desc(Last_Updated))  %>% mutate(new_cases = CaseCount - lead(CaseCount)) %>% slice(1:14)
  cases_7 <- cases_14 %>% filter(Last_Updated > (max(Last_Updated) - 7)) %>% ungroup() %>% select(new_cases) %>% sum()# / population
  cases_previous_week <- cases_14 %>% filter(Last_Updated < (max(Last_Updated)-6)) %>% ungroup() %>% select(new_cases) %>% sum() # / population
  change <- cases_7 - cases_previous_week
  #################
  # PTs <- data$prname[1:13]
  # ii <- which(PTs == i)
  
  HZ_table_values[ii,1:4] <- data.frame(cases_14$Province[1], i, cases_7, change)
  ii <- ii +1 
}

HZ_info <- read.csv(url("https://opendata.arcgis.com/api/v3/datasets/e5403793c5654affac0942432783365a_0/downloads/data?format=csv&spatialRefId=4326"))

HZ_info <- HZ_info %>% select(HR_UID, ENGNAME, TotalPop2019) %>% distinct()
HZ_info$HR_UID <- as.character(HZ_info$HR_UID)

names(HZ_table_values)[2] <- "HR_UID"

HZ_table_values2 <- merge(HZ_table_values, HZ_info, by="HR_UID")
HZ_table_values2$TotalPop2019 <- HZ_table_values2$TotalPop2019 / 100000
HZ_table_values2$cases_7 <- round(HZ_table_values2$cases_7 / HZ_table_values2$TotalPop2019, digits = 1)
HZ_table_values2$change <- round(HZ_table_values2$change / HZ_table_values2$TotalPop2019, digits = 1)

##### match codes to get names and populations. 

```


```{css, echo=FALSE}
.bar-cell {
  display: flex;
  align-items: center;
  font-size: 10px;
}
.number {
  font-size: 14px;
  white-space: pre;
}
.bar-chart {
  flex-grow: 1;
  margin-left: 6px;
  height: 22px;
  font-size: 10px;
}
.bar {
  height: 100%;
}

.table {
  margin: 0 auto;
  width: 675px;
}
.tableTitle {
  margin: 6px 0;
  font-size: 14px;
  font-family: 'Econ Sans Cnd'
}
.tableTitle h2 {
  font-size: 24px;
  font-weight: bold;
  font-family: 'Econ Sans Cnd'
}
.tableTitle p {
  font-size: 16px;
  font-weight: 500;
}
```

```{r,echo=FALSE, fig.height=4, fig.width=11.2, fig.fullwidth = FALSE, message=FALSE, warning=FALSE}

library(plotly)
library(tidyverse)
library(htmlwidgets)

pfw
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# need to change accents 


QC_names <- HZ_table_values2$ENGNAME[HZ_table_values2$cases_14.Province.1. == "QC"]

# library(rvest)
# 
# link <- "https://www.msss.gouv.qc.ca/en/reseau/regions-sociosanitaires-du-quebec/"
# 
# page <- read_html(link)
# new_QC_names <- page %>% html_nodes("h3") %>% html_text()

# remove html from names
#new_QC_names <- gsub('.*\\. ',"",new_QC_names))


new_QC_names <- c ("Bas-Saint-Laurent", "Saguenay–Lac-Saint-Jean", "Capitale-Nationale", "Mauricie-et-Centre-du-Québec", "Estrie", "Montréal", "Outaouais", "Abitibi-Témiscamingue", "Côte-Nord", "Nord-du-Québec", "Gaspésie–Îles-de-la-Madeleine", "Chaudière-Appalaches", "Laval", "Lanaudière", "Laurentides", "Montérégie", "Nunavik", "Terres-Cries-de-la-Baie-James")


names_pair <- data.frame(QC_names, new_QC_names)

#######################


HZ_table_values2$ENGNAME[HZ_table_values2$cases_14.Province.1. == "QC"] <- new_QC_names[match(HZ_table_values2$ENGNAME[HZ_table_values2$cases_14.Province.1. == "QC"], QC_names)]
  
  
################  match province names

colnames(HZ_table_values2)[2] <- "PT"
pt_names <- unique(HZ_table_values2$PT)

new_pt_names <- c("Newfoundland and Labrador", "Prince Edward Island", "Nova Scotia", "New Brunswick", "Québec", "Ontario", "Manitoba", "Saskatchewan", "Alberta", "British Columbia", "The Yukon", "Northwest Territories", "Nunavut")


#######################

HZ_table_values2$PT <- new_pt_names[match(HZ_table_values2$PT, pt_names)]

  
  

```



```{r, echo=FALSE}


append_date_suffix <- function(dates) {
  suff <- case_when(
    dates %in% c(11, 12, 13) ~ "th",
    dates %% 10 == 1 ~ 'st',
    dates %% 10 == 2 ~ 'nd',
    dates %% 10 == 3 ~ 'rd',
    TRUE ~ "th"
  )
  paste0(dates, suff)
}

bar_style <- function(width = 1, fill = "#e6e6e6", height = "100%", align = c("left", "right"), color = NULL) {
  align <- match.arg(align)
  if (align == "left") {
    position <- paste0(width * 100, "%")
    image <- sprintf("linear-gradient(90deg, %1$s %2$s, transparent %2$s)", fill, position)
  } else {
    position <- paste0(100 - width * 100, "%")
    image <- sprintf("linear-gradient(90deg, transparent %1$s, %2$s %1$s)", position, fill)
  }
  list(
    backgroundImage = image,
    backgroundSize = paste("100%", height),
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center",
    color = color
  )
}


# bar_chart_pos_neg <- function(width = 1, fill = "#e6e6e6", height = "75%", align = c("left", "right"), pos_fill = "#ff121a", neg_fill = "#02aa5c") {
#   neg_chart <- div(style = list(flex = "1 1 0"))
#   pos_chart <- div(style = list(flex = "1 1 0"))
#   width <- paste0(abs(value / max(max(HZ_table_values2$cases_7))) * 100, "%")
#   
#   if (value < 0) {
#     bar <- div(style = list(marginLeft = "8px", background = neg_fill, width = width, height = height))
#     chart <- div(style = list(display = "flex", alignItems = "center", justifyContent = "flex-end"), label, bar)
#     neg_chart <- tagAppendChild(neg_chart, chart)
#   } else {
#     bar <- div(style = list(marginRight = "8px", background = pos_fill, width = width, height = "75%"))
#     chart <- div(style = list(display = "flex", alignItems = "center"),  bar, label)
#     pos_chart <- tagAppendChild(pos_chart, chart)
#   }
#   
#   div(style = list(display = "flex"), neg_chart, pos_chart)
# }



bar_style_pos_neg <- function(width = 1, pos_fill = "lightpink", neg_fill = "lightgreen") {
  # Split the background into 2 halves for negative and positive bars.
  # For positive bars, draw the bar from 50% to 50% + width
  # For negative bars, draw the bar from 50% + -width to 50%
  position <- paste0((0.5 + width / 2) * 100, "%")
  if (width >= 0) {
    background <- sprintf("linear-gradient(90deg, transparent 50%%, %2$s 50%%, %2$s %1$s, transparent %1$s)", position, pos_fill)
  } else {
    background <- sprintf("linear-gradient(90deg, transparent %1$s, %2$s %1$s, %2$s 50%%, transparent 50%%)", position, neg_fill)
  }

  list(
    background = background,
    backgroundSize = "98% 88%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
}


# bar_chart_pos_neg <- function(label, value, max_value = 1, height = "16px",
#                               pos_fill = "#ff121a", neg_fill = "#02aa5c") {
#   neg_chart <- div(style = list(flex = "1 1 0"))
#   pos_chart <- div(style = list(flex = "1 1 0"))
#   width <- paste0(abs(value / max(max(HZ_table_values2$cases_7))) * 100, "%")
#   
#   if (value < 0) {
#     bar <- div(style = list(marginLeft = "8px", background = neg_fill, width = width, height = height))
#     chart <- div(style = list(display = "flex", alignItems = "center", justifyContent = "flex-end"), label, bar)
#     neg_chart <- tagAppendChild(neg_chart, chart)
#   } else {
#     bar <- div(style = list(marginRight = "8px", background = pos_fill, width = width, height = height))
#     chart <- div(style = list(display = "flex", alignItems = "center"),  bar, label)
#     pos_chart <- tagAppendChild(pos_chart, chart)
#   }
#   
#   div(style = list(display = "flex"), neg_chart, pos_chart)
# }


########################### The table

HZ_table_values2 <- HZ_table_values2 %>% select(2, ENGNAME, cases_7, change)

table <- reactable(HZ_table_values2, defaultSortOrder = 'desc',
  defaultSorted = 'cases_7', height = 500,
  theme = reactableTheme(
    borderColor = "#dfe2e5",
    stripedColor = "#f6f8fa"),
  showSortIcon = TRUE,
  compact = TRUE, filterable = TRUE, pagination = FALSE,  defaultColDef = colDef(
    ### define header styling
    headerStyle = list(
      textAlign = "left",
      fontSize = "16px",
      lineHeight = "16px",
      textTransform = "uppercase",
      color = "#0c0c0c",
      fontWeight = "500",
      borderBottom = "2px solid #e9edf0",
      paddingBottom = "2px",
      verticalAlign = "bottom",
      fontFamily = font_es,
       footer = function(values, name) htmltools::div(name, style = list(fontWeight = 600))
    ),
    ### define default column styling
    style = list(
      fontFamily = font_es,
      fontSize = "14px",
      verticalAlign = "center",
      align = "left"
    )
  ),  columns = list(
    ENGNAME = colDef(
      name = "Region",
      maxWidth = 200
      #style = list(fontsize = "6px")
    ), 
    PT = colDef(
      name = "Province / Territory",
      maxWidth = 150
      #style = list( fontsize = 6)
    ),
    cases_7 = colDef(
      name = "Cases over past 7 day / 100,000",
      align = "left",
      #fontSize = "14px",
      style = function(value) {
        bar_style(width = value / max(HZ_table_values2$cases_7))
      },
      maxWidth = 250,
      ), 
    change = colDef(
      name = "Change in case rate from previous week",
       align = "center",
        #fontSize = "14px",
      # style = list(fontFamily = font_es,
      # fontWeight = "400", fontsize = "13px"),
      maxWidth = 250,
      style = function(value) {
      bar_style_pos_neg(width = value / max(abs(HZ_table_values2$cases_7)))
     # bar_chart_pos_neg(label, value, height)
    })
      ))


div(class = "tableTitle",
    div(
      class = "title",
      h2("Interactive Table Showing Regional COVID-19 Cases and Week to Week Changes"),
      p(
        paste0("Updated on ", format(Sys.Date(),"%B "),
          append_date_suffix(
            as.numeric(format(Sys.Date(), "%d"))
            ), " ",
          strftime(Sys.time(), format = "%H:%M"), " ET.", " Click on header title to sort table, or search for a region using the search tab."
        )
      ),
    ),
    table)
 
      
      
```


